An evaluation framework for trajectory inference methods

### 2017-12-06 analysis/paper/paper.Rmd
traj types

Dozens[b] of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`). Initially, most trajectory inference was focused on linear trajectories, with one start and one end point. However, the prospect of branching trajectories was already described early [@trapnell_dynamics_2014], and about half of the current methods can handle one or more bifurcations in the trajectory (`r ref("sfig", "methods_over_time", "b")`). Some methods can already handle more complex, cyclical and branching behaviour, and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].

Although[c] trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction or clustering and (iii) ordering the cells along the simplified representation. These components are frequently interchangeable. Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used or required by the method

@@

```{r, fig.width = 12, fig.height = 9}
method_characteristics <- readRDS(figure_file("method_characteristics.rds", experiment_id="method_characteristics"))

print(method_characteristics)
```

@@

### Evaluation overview
![evaluation_overview](../../analysis/figures/evaluation_overview.png)
`r ref("fig", "evaluation_overview", anchor=TRUE)` Overview of the evaluation

Our evaluation was structured in three main phases. In the first phase, we used toy data as a positive control for each method, assessing whether it can detect the correct trajectories in very simplistic data. Next, we generated synthetic data from a realistic model of gene regulation, combined with a simulation of the technical variation induced by single-cell RNA sequencing. We used this synthetic data in a parameter optimization procedure, where we used cross-validation between datasets to avoid overfitting on particular data set structures. Finally, in the third phase, we used the optimized parameters to score the performance of each method on real data. 
### Score metrics and parameter optimisation

@@

## Trajectory inference methods
## Toy data generation
## Synthetic data generation
Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@marbach_wisdom_2012], but required several additions.

>>>>

Dozens[b] of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`). Initially, most trajectory inference was focused on linear trajectories, with one start and one end point. However, the prospect of branching trajectories was already described early [@trapnell_dynamics_2014], and about half of the current methods can handle one or more bifurcations in the trajectory (`r ref("sfig", "methods_over_time", "b")`). Some methods can already handle more complex, cyclical and branching behaviour [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].

Although[c] trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction or clustering and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016]. Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used or required by the method, which can both bias the trajectory to current knowledge but also guide the method towards choosing the right trajectory among many others. Furthermore, with the ongoing strong decreases in single-cell -omics technologies, it becomes

@@

![](`r figure_file("figure_methods.svg", experiment_id="method_characteristics")`)

@@

### Evaluation overview

![evaluation_overview](../../analysis/figures/evaluation_overview.png)
`r ref("fig", "evaluation_overview", anchor=TRUE)` Overview of the evaluation

Our evaluation was structured in three main phases. In the first phase, we used toy data as a positive control for each method, assessing whether it can detect the correct trajectories in very simplistic data. Next, we generated synthetic data from a realistic model of gene regulation, combined with a simulation of the technical variation induced by single-cell RNA sequencing. We used this synthetic data in a parameter optimization procedure, where we used cross-validation between datasets to avoid overfitting on particular data set structures. Finally, in the third phase, we used the optimized parameters to score the performance of each method on real data. We split the real datasets in two groups based on the type of reference trajectory available. The reference in datasets with a silver standard were defined by the author’s based on the expression data itself, while the reference for gold standard was defined using external information, such as fluorescent sorting or time series.
### Score metrics and parameter optimisation

@@

## Trajectory inference methods
## Toy data generation
## Synthetic data generation

Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment. At every step, great care was taking to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable.

### Network generation

One of the main process involved in cellular dynamic processes is through gene regulation, where regulatory cascades and feedback loops lead to . The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo. One driver of bifurcations seem to be mutual antognism, where two genes  or gene modules strongly inhibit eachother, forcing one of the two to become inactive. The choice of which gene(s) survive can be both due to stochasticity [ref] or driven by environmental cues.Of Modules? [Th17]
